---
title: "Project"
format: html
editor: visual
---

Load in the data

```{r}
file <- read.csv("tempch4.txt")

```

Loading the packages

```{r}
library(tidyverse)
```

Check the data

```{r}
glimpse(file)
```

Split the datetime column to filter with different time variable

```{r}
data2 <- separate(file, col = datetime, into = c("Date", "Time"), sep = " ")
data <- separate(data2, col = Date, into = c("Year", "Month", "Day"), sep = "-")
```

Basic Descriptive statistic

```{r}
data |> 
summary()
```

Generating the variance and Std

```{r}
data |> 
  summarise(Number = n(), AvgT = mean(temp_wew), 
            AvgH = mean(rh,  na.rm = TRUE), StdT = sd(temp_wew), StdH = sd(rh, na.rm = TRUE))
```

```{r}
file["month"] <- data[["Month"]]
```

View the distribution of the two variables

```{r}
# Distributio of Temperature
Temperature <- file$temp_wew
hist(Temperature)
```

Almost Normally distributed, but there is appearance of almost two peaks.

```{r}
# Distributio of Humidity
rtv_humidity <- file$rh
hist(rtv_humidity)
```

Plot the temperature over the whole period, BY HOUR

```{r}
file |> 
  ggplot(aes(x= datetime, y = temp_wew, colour = month)) +
  geom_point()+
  geom_hline(yintercept = 23)
```

The Colors represents different months, starting from Feb to Dec. There is trend showing a bell distribution. However, some months (summer) appear to experience higher temperature above 23, while some are below.

For better visualization, The monthly average temperature over the year, signify a bell shape. Plotting the mean of each months over the year.

```{r}
  data |> 
  group_by(Month) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
    ggplot(aes(x = Month,y =AverageT, group = 1)) +
  geom_line(colour = "blue") +
  geom_point() 
```

```{r}
file |> 
  ggplot(aes(x= datetime, y = rh, colour = month)) +
  geom_point()
```
For Humidity, the pattern is different.. higher humidity towards the end of the year and higher variance at the mid months.

```{r}
  data |> 
  group_by(Month) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
    ggplot(aes(x = Month,y =AverageH, group = 1)) +
  geom_line(colour = "green") +
  geom_point() 
```

Plotting the mean of the distribution, it shows increasing trend of humidity.

HOURLY DISTRIBUTION DURING THE DAY

```{r}
data |> 
  group_by(Time) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
  ggplot(aes(x = Time,y =AverageT, group = 1)) +
  geom_line(colour = "blue") +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The average hourly temperature for the whole year, shows rising and falling pattern

```{r}
data |> 
  filter(Month == "07") |> 
  group_by(Time) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
  ggplot(aes(x = Time,y =AverageT, group = 1)) +
  geom_line(colour = "blue") +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Filtering through different months to see changes in the distribution, the pattern still follow similar trends. Temperature decreasing from early hour(00:00) and to lowest temperature at 06:00 and rise from the hour to the highest at region between 16-19:00 and started to decrease again.

```{r}
data |> 
  group_by(Time) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
  ggplot(aes(x = Time,y =AverageH, group = 1)) +
  geom_line(colour = "red") +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

FOR

```{r}
data |> 
  filter(Month == "09") |> 
  group_by(Time) |> 
  summarise(AverageT = mean(temp_wew), AverageH = mean(rh, na.rm = TRUE)) |> 
  ggplot(aes(x = Time,y =AverageH, group = 1)) +
  geom_line(colour = "red") +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Filtering through the months, the distribution is more stable compare to the temperature pattern of the year.

```{r}
  data2 |> 
  group_by(Date) |> 
  summarise(Tmax = max(temp_wew), Tmin = min(temp_wew)) |> 
  ggplot() +
  geom_point(aes(x= Date,y = Tmax), color = "red") +
  geom_point(aes(x= Date,y = Tmin, group = 1), color = "blue") 
  
```

The above graph represent, daily max and minimum temperature.Both follow the same trends

Daily change (Hmax-Hmin) in temperature over the period.

```{r}
data2 |> 
  group_by(Date) |> 
  summarise(ddT = max(temp_wew)-min(temp_wew), ddH = max(rh)-min(rh)) |> 
  ggplot(aes(x= Date, y = ddT, group = 1)) +
  geom_point()
```

```{r}
data2 |> 
  group_by(Date) |> 
  summarise(Hmax = max(rh), Hmin = min(rh)) |> 
  ggplot() +
  geom_point(aes(x= Date,y = Hmax), color = "red") +
  geom_point(aes(x= Date,y = Hmin, group = 1), color = "blue") 
```

Daily change (Tmax-Tmin) in humidi over the period.

```{r}
data2 |> 
  group_by(Date) |> 
  summarise(ddT = max(temp_wew)-min(temp_wew), ddH = max(rh)-min(rh)) |> 
  ggplot(aes(x= Date, y = ddH, group = 1)) +
  geom_point()
```
